Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-17 submission/manuscript.Rmd
discussion motifs

### ChIPseq motifs and transcription factors

Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which different classes of germline genes can be misexpressed in _Kdm5c_-KO cells. Transcription of germline genes in the mature _Kdm5c_-KO brain is likely due to loss of KDM5c-mediated repression during embryogenesis, given that KDM5C was bound to their promoters in EpiLCs, is not recruited to germline genes in neurons, and germline transcription is not repressed when KDM5C in reintroduced in neurons[@scandagliaLossKdm5cCauses2017],

However, it is unclear how KDM5C is recruited to germline genes given that KDM5C itself does not contain domains for sequence-specific binding.

 In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound promoters in EpiLCs, only about one third of KDM5C-bound germline genes contained their consesus sequence at their promoter. Thus, other, unknown factors likely facilitate KDM5C's recruitment to germline genes.

We also found germline genes can become activated in _Kdm5c_-KO cells indepedent of direct loss of KDM5C suppression, given that KDM5C was only bound to about a third of germline-enriched DEGs. One notable EpiLC DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells to promote meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. However, retinoic acid can only activate _Stra8_ expression when DAZL is present. Unlike _Stra8_, _Dazl_ is a direct target of KDM5C in EpiLCs and is transcribed _Kdm5c_-KO EpiLCs. Expression of indirect DEGs like _Stra8_ indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells through downstream activation by other ectopic germline programs. In support of this, germline genes unbound by KDM5C were significantly enriched for RFX transcription factors. RFX transcription factors bind to X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote the transcription of cilia and flagellar genes[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009]. One of the top RFX members enriched included RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016] whose mRNA is also expressed in _Kdm5c_-KO EpiLCs. Intruigingly, _Kdm5c_-KO EpiLCs primarily expressed germline genes involved in early germ cell development and meiosis while the mature _Kdm5c_-KO brain expressed late-stage spermatogenesis genes, including known RFX2 targets. Altogether, these data suggest once activated due to loss of direct KDM5C repression, ectopic germline programs can loosely mimic germline development and progress in the background of _Kdm5c_-KO somatic development.

>>>>

Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which different classes of germline genes can be misexpressed in _Kdm5c_-KO cells. Transcription of germline genes in the mature _Kdm5c_-KO brain is likely due to loss of KDM5c-mediated repression during embryogenesis, given that KDM5C was bound to their promoters in EpiLCs, is not recruited to germline genes in neurons, and germline transcription is not repressed when KDM5C in reintroduced in neurons[@scandagliaLossKdm5cCauses2017], 
It is unclear how KDM5C is recruited to germline genes, given that KDM5C itself does not contain domains for sequence-specific binding. In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound germline promoters in EpiLCs, only about one-third of promtoers contained their consesus sequence. Additionally, germline genes like _Stra8_ can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters. _Stra8_ is a meiotic transcription factor that is typically turned on in the presence of retinoic acid signaling and DAZL[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006], which is a direct KDM5C target. Germline genes unbound by KDM5C were primarily late-stage spermatogenesis genes and significantly enriched for RFX transcription factors that promote the transcription of cilia and flagellar genes[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009]. One of the top RFX members enriched included RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016] whose mRNA is also expressed in _Kdm5c_-KO EpiLCs. Altogether, these data suggest that once germline genes are activated due to loss of direct KDM5C repression, they can activate other ectopic germline programs downstream that loosely mimic germline development.
